Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic

### 2024-11-25 manuscript/manuscript.qmd
Address R&R issue 1

What these findings point to is that democratic backsliding is a complex phenomenon---states experiencing backsliding impose discriminatory policies and abusive behaviors, but part of their institutional embeddedness remains invested, at least in part, in communicating some of these problems and maintaining connections with the international community around human rights. It is important to note that this study maps behavior that occurred but it does not fully theorize or test why backsliders were motivated to care about and submit derogations. We encourage future research on democratic backsliding and human rights to more fully examine backslider consideration of reputation, foreign policy during backsliding, and specifically international human rights law behavior during backsliding. Research might also track how continued engagement with international human rights law might shape, or mitigate, backsliding itself. Future research also would benefit the study of whether and how derogations might be used after a crisis has ended. Do states continue using crisis justification after a local/regional/global crisis is largely over? Overall, our findings contribute to the understanding of how neither derogation filing nor backsliding alone depict a complete picture of human rights and legal behaviors during the pandemic.

>>>>

What these findings point to is that democratic backsliding is a complex phenomenon---states experiencing backsliding impose discriminatory policies and abusive behaviors, but part of their institutional embeddedness remains invested, at least in part, in communicating some of these problems and maintaining connections with the international community around human rights. It is important to note that this study maps behavior that occurred but it does not fully theorize or test why backsliders were motivated to care about and submit derogations. We encourage future research on democratic backsliding and human rights to more fully examine backslider consideration of reputation, foreign policy during backsliding, and specifically international human rights law behavior during backsliding. Research might also track how continued engagement with international human rights law might shape, or mitigate, backsliding itself. Future research also would benefit the study of whether and how derogations might be used after a crisis has ended. Most states did not formally derogate, but nearly all states implemented similar emergency policies. States that did not rely on derogations implemented emergency policies by stretching domestic norms and laws to accommodate expanded government policies. Do states continue using crisis justification after a local/regional/global crisis is largely over, and does the formality of derogation influence the probability of returning to pre-crisis policies? Newer research hints that derogations were associated with a reduced probability of human rights abuses early in the pandemic and that states that filed derogations did not use temporary measures as a carte blanche "pandemic pass" for additional repression [@ChaudhryComstockHeiss:2024]. However, it is too early to tell if formal emergency declarations will encourage or prevent a return to pre-pandemic policies. Overall, our findings contribute to the understanding of how neither derogation filing nor backsliding alone depict a complete picture of human rights and legal behaviors during the pandemic.
